---
title: "Getting started with simulating data in R: some helpful functions and how to use them"
author: "Ariel Muldoon"
date: "August 28, 2018"
output: github_document
---

# Overview

Here's what we'll talk about today:

1. Simulate quantitative variables with `rnorm()` and `runif()`
2. Generate character variables that represent groups via `rep()`.  
3. Use `replicate()` to repeat the data simulation process many times

# Generating random numbers

An easy way to generate numeric data is to pull random numbers from some distribution.  The functions that do this in R always start with the letter `r` (for "random").

## `rnorm()` to generate random numbers from the normal distribution

Pull 5 random numbers from a standard normal distribution.

```{r}
rnorm(5)
```

### Writing out arguments for clearer code

Using the defaults makes for quick coding but does not make the parameters of the generating distribution clear.

```{r}
rnorm(n = 5, mean = 0, sd = 1)
```

### Setting the random seed for reproducible random numbers

To reproduce the random numbers, set the seed via `set.seed()`.

Set the seed and generate 5 numbers.

```{r}
set.seed(16)
rnorm(n = 5, mean = 0, sd = 1)
```

Reset the seed and we get the same 5 numbers.

```{r}
set.seed(16)
rnorm(n = 5, mean = 0, sd = 1)
```

### Change parameters in `rnorm()`

We can pull from different normal distributions by changing the parameters.

First, using a mean of 0 and standard deviation of 2 (so a variance of 4).

```{r}
rnorm(n = 5, mean = 0, sd = 2)
```

Using a large mean and relatively small standard deviation can give values that are strictly positive.

```{r}
rnorm(n = 5, mean = 50, sd = 20)
```

### Using vectors of values for the parameter arguments

Both `mean` and `sd` will take vectors of values.

Let's pull 10 values from distributions with different means but the same standard deviation.

```{r}
rnorm(n = 10, mean = c(0, 5, 20), sd = 1)
```

We could pass a vector to `sd`, as well, but not `n`.  The `n` argument uses the length of the vector to indicate the number of values desired.  

```{r}
rnorm(n = c(2, 10, 10), mean = c(0, 5, 20), sd = c(1, 5, 20) )
```

## Example of using the simulated numbers from `rnorm()`

Exploring how related two unrelated vectors can appear.

```{r}
x = rnorm(n = 10, mean = 0, sd = 1)
y = rnorm(n = 10, mean = 0, sd = 1)
plot(y ~ x)
```

## `runif()` pulls from the uniform distribution

I like `runif()` to to generate continuous data within a set range.

The default is numbers between 0 and 1.

```{r}
runif(n = 5, min = 0, max = 1)
```

But we can do any range.

```{r}
runif(n = 5, min = 50, max = 100)
```

## Example of using the simulated numbers from `runif()`

I like `runif()` to demonstrate how the relative size of the explanatory variable affects the estimated coefficient.

Make a response variable via `rnorm()` and then generate an explanatory variable between 1 and 2.    

```{r}
set.seed(16)
y = rnorm(n = 100, mean = 0, sd = 1)
x1 = runif(n = 100, min = 1, max = 2)
head(x1)
```

Then make an explanatory variable between 200 and 300.

```{r}
x2 = runif(n = 100, min = 200, max = 300)
head(x2)
```

We'll use the data in a regression model fit via `lm()`, making note of the coefficients.

```{r}
lm(y ~ x1 + x2)
```

# Generate character vectors with `rep()`

Simulations involve categorical variables, as well, that often need to be repeated in a pattern.

## Using `letters` and `LETTERS`

These are *built in constants* in R, and convenient for making a simple character vectors.

The first two lowercase letters.

```{r}
letters[1:2]
```

The last 17 uppercase letters.

```{r}
LETTERS[10:26]
```

## Repeat each element of a vector with `each`

With `each` we repeat each unique character in the vector some number of times.

```{r}
rep(letters[1:2], each = 3)
```

## Repeat a whole vector with the `times` argument

Repeating is different with `times`.

```{r}
rep(letters[1:2], times = 3)
```

## Set the output vector length with the `length.out` argument

Using `length.out` is similar to `times` but the groups can be imbalanced.

```{r}
rep(letters[1:2], length.out = 5)
```

## Repeat each element a different number of `times`

We can get unbalanced data with `times` if we use a vector for the argument.

```{r}
rep(letters[1:2], times = c(2, 4) )
```

## Combining `each` with `times`

When using `times` this way it will only take a single value and not a vector.

```{r}
rep(letters[1:2], each = 2, times = 3)
```


## Combining `each` with `length.out`

This is another way to impart imbalance.

```{r}
rep(letters[1:2], each = 2, length.out = 7)
```

Note you can't use `length.out` and `times` together (if you try, `length.out` will be given priority and `times` ignored).

# Creating datasets with quantiative and categorical variables

## Simulate data with no differences among two groups

We want to simulate a two level grouping variable and a "response" variable where there are no differences among the two groups.

```{r}
group = rep(letters[1:2], each = 3)
response = rnorm(n = 6, mean = 0, sd = 1)
data.frame(group,
           response)
```

We don't have to make each variable separately before putting in a data.frame. 

```{r}
data.frame(group = rep(letters[1:2], each = 3),
           response = rnorm(n = 6, mean = 0, sd = 1) )
```

Now let's add another categorical variable to this dataset that's *crossed* with the first.  

The new factor will have three values.  

```{r}
LETTERS[3:5]
```

Remember the `group` factor is repeated elementwise.

```{r}
rep(letters[1:2], each = 3)
```

So what argument do we use for the new variable?

```{r, eval = FALSE}
rep(LETTERS[3:5], ?)
```



To repeat the whole vector twice can use the `times` argument or `length.out = 6`.

You can see that every level of this new variable occurs with every level of `group`.

```{r}
data.frame(group = rep(letters[1:2], each = 3),
           factor = rep(LETTERS[3:5], times = 2),
           response = rnorm(n = 6, mean = 0, sd = 1) )
```

What if we tried to use `each` instead?

```{r}
data.frame(group = rep(letters[1:2], each = 3),
           factor = rep(LETTERS[3:5], each = 2),
           response = rnorm(n = 6, mean = 0, sd = 1) )
```

## Simulate data with a difference in means among two groups

We can use a vector for `mean` in `rnorm()` for this.

```{r}
response = rnorm(n = 6, mean = c(5, 10), sd = 1)
response
```


How do we get the `group` pattern correct?

```{r, eval = FALSE}
group = rep(letters[1:2], ?)
```





We need `times` or `length.out` to repeat the whole vector to match the output of `rnorm()`.

```{r}
group = rep(letters[1:2], length.out = 6)
group
```

Getting the order correct is one reason to build vectors separately before binding them into a data.frame.

```{r}
data.frame(group,
            response)
```


# Repeatedly simulating data with `replicate()`

The `replicate()` function is a real workhorse when making repeated simulations as it is for *repeated evaluation of an expression (which will usually involve random number generation)*.

It takes three arguments:

* `n`, which is the number of replications to perform.  This is to set the number of repeated runs we want.  
* `expr`, the expression that should be run repeatedly.  This is often a function.  
* `simplify`, which controls the type of output the results of `expr` are saved into.  Use `simplify = FALSE` to get output saved into a list instead of in an array.  


## Simple example of `replicate()`

Generate 5 values from a standard normal distribution 3 times.

```{r}
set.seed(16)
replicate(n = 3, 
          expr = rnorm(n = 5, mean = 0, sd = 1), 
          simplify = FALSE )
```

Without `simplify = FALSE` we get a matrix.

```{r}
set.seed(16)
replicate(n = 3, 
          expr = rnorm(n = 5, mean = 0, sd = 1) )
```


## An equivalent `for()` loop example

The same thing can be done with a `for()` loop, which I've found can be easier code to follow for R beginners.

```{r}
set.seed(16)
list1 = list() # Make an empty list to save output in
for (i in 1:3) { # Indicate number of iterations with "i"
    list1[[i]] = rnorm(n = 5, mean = 0, sd = 1) # Save output in list for each iteration
}
list1
```

## Using `replicate()` to repeatedly make a dataset

```{r}
simlist = replicate(n = 3, 
          expr = data.frame(group = rep(letters[1:2], each = 3),
                     response = rnorm(n = 6, mean = 0, sd = 1) ),
          simplify = FALSE)
```

We can see this result is a list of three data.frames.

```{r}
str(simlist)
```

Here is the first of the three.

```{r}
simlist[[1]]
```

# What's the next step?

By saving our generated variables or data.frames into a list we've made it so we can loop via list looping functions like `lapply()` or `purrr::map()`.  

You can see a few examples of `replicate()` followed by `map()` in my blog post [A closer look at replicate() and purrr::map() for simulations](https://aosmith.rbind.io/2018/06/05/a-closer-look-at-replicate-and-purrr/).
